UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-08-29 ms/main.tex
add some citations
minor tweaks to text, final citations

Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in fig~\ref{fig:supp_megamontage_pc10_50}. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations. UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figs, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

Scripts for all tests and plotting functions can be found on \url{https://github.com/diazale/gt-dimred}. A demo version using freely available 1KGP data is available at \url{https://github.com/diazale/1KGP_dimred}. PCA and standard t-SNE were done with Scikit-learn\cite{scikit-learn}. UMAP was performed using a Python implementation\cite{2018arXivUMAP}. Statistical testing was done in SciPy\cite{scipy} and StatsModels\cite{seabold2010statsmodels}.

>>>>

Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in fig~\ref{fig:supp_megamontage_pc10_50}. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations. UMAP is sensitive to sample sizes and spends more visual space on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figs, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

Scripts for all tests and plotting functions can be found on \url{https://github.com/diazale/gt-dimred} with a command line script for UMAP available at \url{https://github.com/diazale/gt-dimred/scripts/general_umap_script.py}. A demo version using freely available 1KGP data is available at \url{https://github.com/diazale/1KGP_dimred}. PCA and standard t-SNE were done with Scikit-learn\cite{scikit-learn}. UMAP was performed using a Python implementation\cite{2018arXivUMAP}. Statistical testing was done in SciPy\cite{scipy}, StatsModels\cite{seabold2010statsmodels}, and R\cite{r2013}. Visualizations were created with Matplotlib\cite{Hunter2007} and ggplot2\cite{ggplot2016}, and maps were created with rgdal\cite{rgdal2019}.
